\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in \proglang{R}

### 2014-11-25 papers/jss/article.Rnw
Add acknowledgements for Tim Hesterberg and two anonymous referees.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/rprotobuf@909 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Updates to section 7:
Add a better transition to start section 7 reminding the user what the
application in section 6 was about and how/why this one is different.

Remove several duplicate sentences about the basics of protocol buffer
.proto files and such which are explained earlier in the paper.

Remove a few sentences that provide unnecessary level of detail about
OpenCPU.  In this section it is an example web service and

As described earlier, the primary application of Protocol Buffers is data
interchange in the context of inter-system communications.  Network protocols
such as HTTP provide mechanisms for client-server communication, i.e., how to
initiate requests, authenticate, send messages, etc.  However, network
protocols generally do not regulate the \emph{content} of messages: they
allow transfer of any media type, such as web pages, static files or
multimedia content.  When designing systems where various components require
exchange of specific data structures, we need something on top of the network
protocol that prescribes how these structures are to be represented in
messages (buffers) on the network. Protocol Buffers solve exactly this
problem by providing a cross-platform method for serializing arbitrary
structures into well defined messages, which can then be exchanged using any
protocol. The descriptors (\code{.proto} files) are used to formally define
the interface of a remote API or network application. Libraries to parse and
generate protobuf messages are available for many programming languages,
making it relatively straightforward to implement clients and servers.

@@

One example of a system that supports Protocol Buffers to interact
with \proglang{R} is OpenCPU \citep{opencpu}. OpenCPU is a framework for embedded statistical 
computation and reproducible research based on \proglang{R} and \LaTeX. It exposes a 
HTTP(S) API to access and manipulate \proglang{R} objects and allows for performing 
remote \proglang{R} function calls. Clients do not need to understand 
or generate any \proglang{R} code: HTTP requests are automatically mapped to 
function calls, and arguments/return values can be posted/retrieved
using several data interchange formats, such as Protocol Buffers.  
OpenCPU uses the \code{serialize\_pb} and \code{unserialize\_pb} functions
from the \pkg{RProtoBuf} package to convert between \proglang{R} objects and protobuf
messages. Therefore, clients need the \code{rexp.proto} descriptor mentioned
earlier to parse and generate protobuf messages when interacting with OpenCPU.

@@

\begin{verbatim}
  GET https://public.opencpu.org/ocpu/library/MASS/data/Animals/pb
\end{verbatim}
The postfix \code{/pb} in the URL tells the server to send this
object in the form of a protobuf message. Alternative formats include 
\code{/json}, \code{/csv}, \code{/rds} and others. If the request
is successful, OpenCPU returns the serialized object with HTTP status 
code 200 and HTTP response header \code{Content-Type: application/x-protobuf}. 
The latter is the conventional MIME type that formally notifies the client to
interpret the response as a protobuf message.

Because both HTTP and Protocol Buffers have libraries available for many 
languages, clients can be implemented in just a few lines of code. Below
is example code for both \proglang{R} and Python that retrieves a data set from \proglang{R} with 
OpenCPU using a protobuf message. In \proglang{R}, we use the HTTP client from 
the \code{httr} package \citep{httr}. In this example we
download a data set which is part of the base \proglang{R} distribution, so we can
verify that the object was transferred without loss of information.

@@

This code suggests a method for exchanging objects between \proglang{R} servers, however this might as 
well be done without Protocol Buffers. The main advantage of using an inter-operable format 
is that we can actually access \proglang{R} objects from within another
programming language. For example, in a very similar fashion we can retrieve the same
data set in a Python client. To parse messages in Python, we first compile the 
\code{rexp.proto} descriptor into a python module using the \code{protoc} compiler:

\begin{verbatim}
  protoc rexp.proto --python_out=.
\end{verbatim}
This generates Python module called \code{rexp\_pb2.py}, containing both the 
descriptor information as well as methods to read and manipulate the \proglang{R} object 
message. In the example below we use the HTTP client from the \code{urllib2}
module.

@@

The example above shows how the \code{HTTP GET} method retrieves a 
resource from OpenCPU, for example an \proglang{R} object. The \code{HTTP POST} 
method on the other hand is used for calling functions and running scripts, 
which is the primary purpose of the framework. As before, the \code{/pb} 
postfix requests to retrieve the output as a protobuf message, in this
case the function return value. However, OpenCPU allows us to supply the
arguments of the function call in the form of protobuf messages as well.
This is a bit more work, because clients needs to both generate messages 
containing \proglang{R} objects to post to the server, as well as retrieve and parse
protobuf messages returned by the server. Using Protocol Buffers to post
function arguments is not required, and for simple (scalar) arguments 
the standard \code{application/x-www-form-urlencoded} format might be sufficient.
However, with Protocol Buffers the client can perform function calls with
more complex arguments such as \proglang{R} vectors or lists. The result is a complete
RPC system to do arbitrary \proglang{R} function calls from within 
any programming language.

The following example \proglang{R} client code performs the remote function call 
\code{stats::rnorm(n=42, mean=100)}. The function arguments (in this
case \code{n} and \code{mean}) as well as the return value (a vector
with 42 random numbers) are transferred using a protobuf message. RPC in
OpenCPU works like the \code{do.call} function in \proglang{R}, hence all arguments
are contained within a list.

@@

The first versions of \pkg{RProtoBuf} were written during 2009 - 2010.
Very significant contributions, both in code and design, were made by
Romain Fran\c{c}ois whose continued influence on design and code is
greatly appreciated. Several features of the package reflect
the design of the \pkg{rJava} package by Simon Urbanek.
The user-defined table mechanism, implemented by Duncan Temple Lang for the
purpose of the \pkg{RObjectTables} package, allows for the dynamic symbol lookup.
Kenton Varda was generous with his time in reviewing code and explaining
obscure Protocol Buffer semantics.  Karl Millar was very
helpful in reviewing code and offering suggestions.  Saptarshi Guha's
work on RHIPE and implementation of a universal message type for \proglang{R}
language objects allowed us to add the \code{serialize_pb} and \code{unserialize_pb}
methods for turning arbitrary R objects into Protocol Buffers without
a specialized pre-defined schema.

>>>>

The previous section described an application where data from a
program written in another language was output to persistent storage
and then read into \proglang{R} for further analysis.  This section
describes another common use case where Protocol Buffers are used as
the interchange format for client-server communication.
Network protocols
such as HTTP provide mechanisms for client-server communication, i.e., how to
initiate requests, authenticate, send messages, etc.  However, network
protocols generally do not regulate the \emph{content} of messages: they
allow transfer of any media type, such as web pages, static files or
multimedia content.  When designing systems where various components require
exchange of specific data structures, we need something on top of the network
protocol that prescribes how these structures are to be represented in
messages (buffers) on the network. Protocol Buffers solve exactly this
problem by providing a cross-platform method for serializing arbitrary
structures into well defined messages, which can then be exchanged using any
protocol.
%The descriptors (\code{.proto} files) are used to formally define
%the interface of a remote API or network application.
%Libraries to parse and
%generate protobuf messages are available for many programming languages,
%making it relatively straightforward to implement clients and servers.

@@

One example of a system that supports Protocol Buffers to interact
with \proglang{R} is OpenCPU \citep{opencpu}. OpenCPU is a framework
for embedded statistical computation and reproducible research based
on \proglang{R} and \LaTeX. It exposes a HTTP(S) API to access and
manipulate \proglang{R} objects and execute remote \proglang{R}
function calls. Clients do not need to understand or generate any
\proglang{R} code: HTTP requests are automatically mapped to function
calls, and arguments/return values can be posted/retrieved using
several data interchange formats, such as Protocol Buffers.  OpenCPU
uses the \code{rexp.proto} descriptor and the \code{serialize\_pb} and
\code{unserialize\_pb} functions described in
Section~\ref{sec:evaluation} to convert between \proglang{R} objects
and protocol buffer messages.

@@

\begin{verbatim}
  GET https://public.opencpu.org/ocpu/library/MASS/data/Animals/pb
\end{verbatim}
The postfix \code{/pb} in the URL tells the server to send this
object in the form of a protocol buffer message.
% Alternative formats include \code{/json}, \code{/csv}, \code{/rds} and others.
If the request
is successful, OpenCPU returns the serialized object with HTTP status
code 200 and HTTP response header \code{Content-Type: application/x-protobuf}.
The latter is the conventional MIME type that formally notifies the client to
interpret the response as a protocol buffer.

Because both HTTP and Protocol Buffers have libraries available for many
languages, clients can be implemented in just a few lines of code. Below
is example code for both \proglang{R} and Python that retrieves an \proglang{R} data set encoded as a protocol buffer message from OpenCPU.
In \proglang{R}, we use the HTTP client from
the \code{httr} package \citep{httr}. In this example we
download a data set which is part of the base \proglang{R} distribution, so we can
verify that the object was transferred without loss of information.

@@

Similarly, to retrieve the same data set in a Python client, we first
compile the \code{rexp.proto} descriptor into a python module
using the \code{protoc} compiler:

\begin{verbatim}
  protoc rexp.proto --python_out=.
\end{verbatim}

This generates Python module called \code{rexp\_pb2.py}, containing
both the descriptor information as well as methods to read and
manipulate the \proglang{R} object message. We use the
HTTP client from the \code{urllib2} module in our example to retrieve the
encoded protocol buffer from the remote server then parse and print it
from Python.

@@

The previous example used a simple \code{HTTP GET} method to retrieve
an \proglang{R} object from a remote service (OpenCPU) encoded as a
protocol buffer.
In many cases simple \code{HTTP GET} methods are insufficient, and a
more complete RPC system may need to create compact protocol buffers
for each request to send to the remote server in addition to parsing
the response protocol buffers.

The OpenCPU framework allows us to do arbitrary \proglang{R} function
calls from within any programming language by encoding the arguments
in the request protocol buffer.  The following example \proglang{R}
client code performs the remote function call \code{stats::rnorm(n=42,
mean=100)}. The function arguments (in this case \code{n} and
\code{mean}) as well as the return value (a vector with 42 random
numbers) are transferred using protocol buffer messages. RPC in OpenCPU
works like the \code{do.call} function in \proglang{R}, hence all
arguments are contained within a list.

@@

The first versions of \pkg{RProtoBuf} were written during 2009 - 2010.
Very significant contributions, both in code and design, were made by
Romain Fran\c{c}ois whose continued influence on design and code is
greatly appreciated. Several features of the package reflect
the design of the \pkg{rJava} package by Simon Urbanek.
The user-defined table mechanism, implemented by Duncan Temple Lang for the
purpose of the \pkg{RObjectTables} package, allows for the dynamic symbol lookup.
Kenton Varda was generous with his time in reviewing code and explaining
obscure Protocol Buffer semantics.  Karl Millar and Tim Hesterberg were very
helpful in reviewing code and offering suggestions.  Saptarshi Guha's
work on RHIPE and implementation of a universal message type for \proglang{R}
language objects allowed us to add the \code{serialize_pb} and \code{unserialize_pb}
methods for turning arbitrary R objects into Protocol Buffers without
a specialized pre-defined schema.  Feedback from two anonymous
referees greatly improved both the presentation of this paper and the
package contents.
